Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-10-27 analysis/paper/paper.tex
round 1 of modifications based on reviewer comments

\usepackage{lineno}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}

Keywords:  dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco 
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants,
whether deliberate or accidental; and (2) passive diffusion as alkaloids
and other compounds are transferred from plasma to saliva, and then into
the oral cavity through the salivary glands in the hours to days
following consumption (Cone \& Huestis, 2007). The relation to plasma is
why there is often a close correlation between presence (not
concentration) of drugs in oral fluid and blood (Cone \& Huestis, 2007;
Milman et al., 2011; Wille et al., 2009). The second pathway allows the
identification of parent compounds that are not consumed orally, as long
as they, or their metabolites, are excreted through saliva. .

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(Jin \& Yip, 2002; White, 1997). This may also allow various alkaloids
of dietary and medicinal origin to become incorporated in dental plaque.
Dental plaque undergoes frequent mineralisation events, ultimately
causing the entrapped alkaloids and their metabolites to become
preserved within the dental calculus. Barring intentional or accidental
removal of the calculus during life, burial, excavation, and
post-excavation cleaning, the alkaloids can then be detected by various
methods to show a record of consumption during life.

@@

The sample consists of 41 individuals from Middenbeemster, a 19th
century rural Dutch site. The village of Middenbeemster and the
surrounding Beemsterpolder was established in the beginning of the 17th
century, when the Beemster lake was drained to create more farmland,
mainly for the cultivation of cole seeds (de Vries 1978). In 1615, a
decision was made to build a church, and construction started in 1618
(Hakvoort 2013). The excavated cemetery is associated with the
Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
(Lemmers et al., 2013). Archival documents are available for those
buried between AD 1829 and 1866, when the majority of individuals were
interred (Palmer et al., 2016). The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the
industrial revolution (Aten et al., 2012; Palmer et al., 2016).

To reduce the number of potentially confounding factors to account for
in the analysis, we preferentially selected males from the middle adult
age category (35-49 years). The sample consists of 27 males, 11 probable
males, 2 probable females, and 1 female
(Figure~\ref{fig-sample-demography}). We selected males due to a higher
occurrence of pipe notches and dental calculus deposits than females
(unpublished observation).

@@

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years).

@@

Pathological conditions and lesions that occur frequently in the
population were included in the analysis. Data were dichotomised to
presence/absence to allow statistical analysis. Osteoarthritis was
considered present in cases where eburnation was visible on one or more
joint surfaces. Vertebral osteophytosis is identified by marginal
lipping and/or osteophyte formation on the margin of the superior and
inferior surfaces of the vertebral body. Cribra orbitalia was diagnosed
based on the presence of pitting on the superior surface of the orbit.
No distinction was made between active or healing lesions. Degenerative
disc disease, or spondylosis, is identified as a large diffuse
depression of the superior and/or inferior surfaces of the vertebral
body (Rogers, 2000). Schmorl's nodes are identified as any cortical
depressions on the surface of the vertebral body. Data on chronic
maxillary sinusitis from Casna et al. (2021) were included in this study
to assess the relationship between upper respiratory diseases with
environmental factors (i.e.~tobacco smoke, caffeine consumption).
Lesions associated with chronic maxillary sinusitis as defined by
Boocock et al. (1995) were recorded for each individual and classified
as ``pitting'', ``spicule-type bone formation'', ``remodeled spicules'',
or ``white pitted bone''. chronic maxillary sinusitis was scored as
absent when the sinus presented smooth surfaces with little or no
associated pitting.

@@

All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
were created using \textbf{ggplot2} (H. Wickham, 2016). Polychoric
correlations were calculated with the \textbf{psych} package (Revelle,
2022).

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
Figure~\ref{fig-auth-plot-batch2}.

@@

\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2.}

@@

\hypertarget{tbl-pearson}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0952}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}@{}}
\caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
dichotomous skeletal lesions and compound concentrations (ng/mg) from
the second batch. Correlations between pairs of dichotomous variables
are removed due to incompatibility with a Pearson correlation. OA =
osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD
= degenerative disc disease; CO = cribra orbitalia; CMS = chronic
maxillary sinusitis; SA = salicylic acid; PN = pipe
notches.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
OA & -0.12 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
VOP & -0.088 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
SN & -0.24 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
DDD & -0.0038 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 &
-0.0086 \\
CO & 0.064 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
CMS & -0.19 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & -0.0045 & -0.22 \\
Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
SA & & & & 0.14 & 0.37 & 0.038 & 0.17 & -0.17 \\
Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.031 \\
PN & & & & & & -0.16 & 0.18 & -0.0068 \\
Theophylline & & & & & & & 0.51 & 0.36 \\
Caffeine & & & & & & & & 0.078 \\
\end{longtable}

Point-biserial correlation was conducted on paired continuous and
dichotomous variables, to see if any relationships exist between
extracted concentrations and other variables. The strongest
point-biserial (Pearson) correlation correlations were a near-perfect
positive correlation between cotinine and nicotine (0.982), and moderate
correlations between theophylline and nicotine (0.432), caffeine and
theophylline (0.507) (Table~\ref{tbl-pearson}).

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.847). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.487), SN and
cotinine (0.559), DDD and calculus (-0.416), CMS and caffeine (0.53),
caries and periodontitis (0.523), periodontitis and VOP (0.487),
periodontitis and age-at-death (0.407), nicotine and SN (0.53), calculus
and DDD (-0.416), age-at-death and theophylline (-0.45), theophylline
and age-at-death (-0.45), caffeine and periodontitis (0.494), cotinine
and CMS (0.427). Remaining correlations were weak or absent
(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

@@

@@

@@

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{lineno}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 56\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, and especially
children and women.
\end{abstract}

Keywords:  dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco 
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through ingestion of alkaloid-containing plants, whether
deliberate or accidental; and (2) passive diffusion as alkaloids and
other compounds are transferred from plasma to saliva, and then
gradually secreted into the oral cavity through the salivary glands in
the hours-to-days following ingestion (Cone \& Huestis, 2007). The
second pathway allows the identification of parent compounds that do not
enter the mouth (e.g.~injection), as long as they, or their metabolites,
are excreted through the saliva, thus eventually entering the oral
cavity.

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(Jin \& Yip, 2002; White, 1997). This may also allow various alkaloids
of dietary and medicinal origin to become incorporated in dental plaque.
Dental plaque undergoes frequent mineralisation events, ultimately
causing the entrapped alkaloids and their metabolites to become
preserved within the dental calculus. Barring intentional or accidental
removal of the calculus during life, burial, excavation, and
post-excavation cleaning, the alkaloids can then be detected by various
methods to show a record of consumption during life. Because drugs may
be transferred from plasma to saliva, there is often a close correlation
between drugs detected in oral fluid and blood, though there are
differences in detected concentrations (Cone \& Huestis, 2007; Milman et
al., 2011; Wille et al., 2009). This was also shown to be true for
dental calculus and blood (Sørensen et al., 2021), making dental
calculus a potentially useful substance for detecting ancient alkaloids
and other dietary compounds.

@@

The sample consists of 41 individuals from Middenbeemster, a 19th
century rural Dutch site. The village of Middenbeemster and the
surrounding Beemsterpolder was established in the beginning of the 17th
century, when the Beemster lake was drained to create more farmland,
mainly for the cultivation of cole seeds (de Vries 1978). In 1615, a
decision was made to build a church, and construction started in 1618
(Hakvoort 2013). The excavated cemetery is associated with the
Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
(Lemmers et al., 2013). Archival documents are available for those
buried between AD 1829 and 1866, when the majority of individuals were
interred. The main occupation of the inhabitants was dairy farming,
consisting largely of manual labour prior to the industrial revolution
(Aten et al., 2012; Palmer et al., 2016).

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

@@

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

@@

Pathological conditions and lesions that occur frequently in the
population were included in the analysis. Data were dichotomised to
presence/absence to allow for statistical analysis. Osteoarthritis was
considered present in cases where eburnation was visible on one or more
joint surfaces. Vertebral osteophytosis is identified by marginal
lipping and/or osteophyte formation on the margin of the superior and
inferior surfaces of the vertebral body. Cribra orbitalia was diagnosed
based on the presence of pitting on the superior surface of the orbit.
No distinction was made between active or healing lesions. Degenerative
disc disease, or spondylosis, is identified as a large diffuse
depression of the superior and/or inferior surfaces of the vertebral
body (Rogers, 2000). Schmorl's nodes are identified as any cortical
depressions on the surface of the vertebral body. Data on chronic
maxillary sinusitis from Casna et al. (2021) were included in this study
to assess the relationship between upper respiratory diseases with
environmental factors (i.e.~tobacco smoke, caffeine consumption).
Lesions associated with chronic maxillary sinusitis as defined by
Boocock et al. (1995) were recorded for each individual and classified
as ``pitting'', ``spicule-type bone formation'', ``remodeled spicules'',
or ``white pitted bone''. chronic maxillary sinusitis was scored as
absent when the sinus presented smooth surfaces with little or no
associated pitting.

@@

All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
were calculated with the \textbf{psych} package (Revelle, 2022).

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
(Figure~\ref{fig-auth-plot-batch2}).

@@

\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2. r = Pearson correlation
coefficient.}

@@

\hypertarget{tbl-pearson}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1413}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0761}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}@{}}
\caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
dichotomous skeletal lesions and compound concentrations (ng/mg) from
the second batch. Correlations between pairs of dichotomous variables
are removed due to incompatibility with a Pearson correlation. Moderate
and strong correlations in \textbf{bold}. OA = osteoarthritis; VOP =
vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc
disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA =
salicylic acid; PN = pipe notches.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
OA & -0.12 & -0.07 & 0.21 & 0.07 & 0.14 & 0.28 & 0 & -0.07 \\
VOP & -0.09 & -0.16 & 0.34 & 0.06 & 0.25 & -0.06 & 0.01 & -0.13 \\
SN & -0.24 & 0.16 & 0.09 & 0.09 & 0.17 & 0.24 & 0.16 & 0.09 \\
DDD & 0 & 0 & 0.19 & -0.39 & -0.08 & 0.31 & 0.06 & -0.01 \\
CO & 0.06 & -0.05 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.06 \\
CMS & -0.19 & 0.28 & 0 & -0.27 & 0.03 & 0.19 & 0.36 & 0.22 \\
Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & 0 & -0.22 \\
Nicotine & & & -0.21 & 0.01 & -0.01 & \textbf{0.43} & 0.14 &
\textbf{0.98} \\
SA & & & & 0.14 & 0.37 & 0.04 & 0.17 & -0.17 \\
Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.03 \\
PN & & & & & & -0.16 & 0.18 & -0.01 \\
Theophylline & & & & & & & \textbf{0.51} & 0.36 \\
Caffeine & & & & & & & & 0.08 \\
\end{longtable}

Point-biserial correlation was conducted on paired continuous and
dichotomous variables, to see if any relationships exist between
extracted concentrations and other variables. The strongest
point-biserial (Pearson) correlation correlations were a near-perfect
positive correlation between cotinine and nicotine (0.98), and moderate
correlations between theophylline and nicotine (0.43), caffeine and
theophylline (0.51) (Table~\ref{tbl-pearson}).

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.85). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
caries and periodontitis (0.52), periodontitis and VOP (0.49),
periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
CMS (0.43). Remaining correlations were weak or absent
(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

@@

\leavevmode\vadjust pre{\hypertarget{ref-huangDecipheringGenetic2023}{}}%
Huang, Y., Shan, Y., Zhang, W., Lee, A. M., Li, F., Stranger, B. E., \&
Huang, R. S. (2023). Deciphering genetic causes for sex differences in
human health through drug metabolism and transporter genes. \emph{Nature
Communications}, \emph{14}(1, 1), 175.
\url{https://doi.org/10.1038/s41467-023-35808-6}

@@

\leavevmode\vadjust pre{\hypertarget{ref-kingCautionaryTales2017}{}}%
King, A., Powis, T. G., Cheong, K. F., \& Gaikwad, N. W. (2017).
Cautionary tales on the identification of caffeinated beverages in
{North America}. \emph{Journal of Archaeological Science}, \emph{85},
30--40. \url{https://doi.org/10.1016/j.jas.2017.06.006}

@@

\leavevmode\vadjust pre{\hypertarget{ref-unoSexAgedependent2017}{}}%
Uno, Y., Takata, R., Kito, G., Yamazaki, H., Nakagawa, K., Nakamura, Y.,
Kamataki, T., \& Katagiri, T. (2017). Sex- and age-dependent gene
expression in human liver: {An} implication for drug-metabolizing
enzymes. \emph{Drug Metabolism and Pharmacokinetics}, \emph{32}(1),
100--107. \url{https://doi.org/10.1016/j.dmpk.2016.10.409}
